---
title: "Namaste National Analyses"
subtitle: "Multivariate changes, at site level"
date: today
format: 
    html:
        toc: true
        toc-location: left
        code-fold: true
        embed-resources: true
    pdf:
        toc: true
        echo: false
warning: false
message: false
error: true
fig-width: 7
fig-height: 5
---

```{r}
library(tidyverse)
library(vegan)
library(ggrepel)
library(ecole)
```


## read and prepare data  

Using the "slopes" by site data frame - changes at the site level.  

```{r}
dat <- read.csv(here::here("data",
                           "compiled",
                           "slopesAndExpl_bySite.csv"))

dat_main <- dat %>% 
    select(Reserve,
           SiteID,
           Total.unvegetated_slope,
           Total.live.veg_slope,
           EIR_slope,
           H.Halophyte_slope,
           Richness_slope,
           SWdiv_slope,
           tide_range_slope:Region,
           condition = NERRs_Landscape_resiliency_condition_sum_quantile,
           SET_change:last_col()
    ) %>% 
    mutate(# fix Geomorphyology
        Geomorphology = case_when(Geomorphology == "Back Barrier" ~ "Back barrier",
                                  .default = Geomorphology),
        # turn character cols into factors
        across(where(is.character), as.factor),
        # turn appropriate columns into ordered factors
        Tidal_Range = factor(Tidal_Range, levels = c("Microtidal", "Mesotidal", "Macrotidal"), ordered = TRUE),
        Salinity_category = factor(Salinity_category,
                                   levels = c("Fresh", "Oligohaline",
                                              "Mesohaline", "Polyhaline"), ordered = TRUE)
    ) %>% 
    filter(!is.na(H.Halophyte_slope))

responses <- dat_main %>% 
    select(Total.unvegetated_slope:SWdiv_slope)
row.names(responses) = paste(dat_main$Reserve, dat_main$SiteID, sep = "-")

responses_scaled <- responses %>% 
    mutate(across(everything(), function(x) scale(x)[,1]))

explanatory <- dat_main[which(!(names(dat_main) %in% names(responses)))]

explanatory_scaled <- explanatory %>% 
    mutate(across(where(is.numeric), 
                  function(x) scale(x)[,1]))

explanatory_scaled_main <- explanatory_scaled %>% 
    select(Geomorphology, Tidal_Range, Salinity_category,
           SLR_last19yrs, NERR_Region, condition,
           SET_minus_SLR_19yrs, proportion_low)
```




# PERMANOVA  

One PERMANOVA for each key factor we care about. Setting this up like a multiple regression means each predictor's effect (and p-value) is influenced by what other predictors are present in the model. We just want to know if, overall, aside from anything else, each of these are different.  

If a PERMANOVA is significant for a factor, then pairwise comparisons will be run: pairwise PERMANOVAs to see which groups are different from which other groups, and SIMPER, which is pairwise between groups to show which pieces of the response matrix are most influencing differences between groups.  

NOTE that SIMPER will run all pairs whether the PERMANOVA showed significance or not, so be sure to check the pairwise PERMANOVA before getting too excited about SIMPER output!  

```{r}
# set up a loop to make a 2nd level header - ## - for each
# key predictor  

# below that, ### overall permanova, then if that's significant,
# ### pairwise permanovas, ### SIMPER  

# might have to write my own function for numeric predictors
# simper definitely won't work on numeric predictors
# permanova_pairwise turns them into factors

key_preds_cat <- c("NERR_Region", "Salinity_category",
                    "Geomorphology", "Tidal_Range")
key_preds_num <- c("condition", "SLR_last19yrs",
                    "SET_minus_SLR_19yrs", "proportion_low")

for(i in seq_along(key_preds_cat)){
    cat("\n\n---------------------------------------------------------- \n")
    cat("---------------------------------------------------------- \n\n")
    cat(paste0("~~~", key_preds_cat[i], "~~~\n\n"))

    explanatory_tmp <- explanatory
    names(explanatory_tmp)[which(names(explanatory_tmp) == key_preds_cat[i])] <- "predictor"
    
    perm <- adonis2(responses_scaled ~ predictor,
                    data = explanatory_tmp,
                    method = "euclidean",
                    na.action = "na.omit"
    )
    
    cat(paste0("PERMANOVA overall for ", key_preds_cat[i], "\n\n"))
    print(perm)
    
    
    if(perm$`Pr(>F)`[1] < 0.05){
        cat(paste0("PERMANOVA for ", key_preds_cat[i], " was significant at the level alpha = 0.05. Pairwise tests follow. \n\n"))
        cat(paste0("~~Pairwise PERMANOVA, ", key_preds_cat[i], "~~ \n\n"))
        perm_pair <- permanova_pairwise(responses_scaled,
                                        grp = explanatory_tmp$predictor,
                                        method = "euclidean")
        
        print(perm_pair)
        
        cat(paste0("\n\n ~~SIMPER, ", key_preds_cat[i], "~~ \n\n"))
        simp <- simper(responses_scaled, group = explanatory_tmp$predictor)
        
        print(summary(simp))
            
            
    }

}
```


# Ordinations  

```{r}
test_bioenv <- bioenv(responses_scaled, explanatory_scaled_main, 
                      method = "spearman",
                      index = "euclidean",
                      metric = "gower")


cat("Explanatory variables used in bioenv: \n", paste("*", names(explanatory_scaled_main)), sep = "\n")

test_bioenv

summary(test_bioenv)
```

## NMDS  

```{r}
veg_mds <- metaMDS(responses_scaled,
                   distance = "euclidean",
                   autotransform = FALSE,
                   trace = FALSE)
cat("2-dimensional stress: ", round(veg_mds$stress, 3))
```

```{r}
data.scores <- as.data.frame(scores(veg_mds))
loadings <- as.data.frame(scores(envfit(veg_mds, responses_scaled), display = "vectors")) *3

# numeric explanatory loadings
expl_numeric <- explanatory_scaled %>% 
    select(SLR_last19yrs,
           condition,
           SET_minus_SLR_19yrs,
           proportion_low)

env_expl <- as.data.frame(scores(envfit(veg_mds, expl_numeric, na.rm = TRUE), display = "vectors")) *5

```



```{r}
# plot all sites
p <- ggplot(data.scores,
            aes(x = NMDS1,
                y = NMDS2)) +
    geom_text(label = rownames(responses_scaled),
              size = rel(3)) +
    geom_segment(data = loadings,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text_repel(data = loadings,
              label = rownames(loadings),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "All sites")
p

# points only
ppts <- ggplot(data.scores,
            aes(x = NMDS1,
                y = NMDS2)) +
    geom_point() +
    geom_segment(data = loadings,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text_repel(data = loadings,
              label = rownames(loadings),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "All sites")
```

## Numeric Loadings overlaid - all  

```{r}
p2 <- ppts +
    geom_segment(data = env_expl,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "blue") +
    geom_text_repel(data = env_expl,
              aes(x = NMDS1, y = NMDS2),
              label = rownames(env_expl),
              col = "blue")

p2
```


## Color by categorical predictors that were important in bioenv  

```{r}
nmds_and_expl <- cbind(data.scores, explanatory_scaled_main)
```

```{r}
plot_nmds2 <- function(expl_color){
    
p <- ggplot(nmds_and_expl,
            aes(x = NMDS1,
                y = NMDS2,
                col = {{expl_color}})) +
    geom_segment(data = loadings,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text_repel(data = loadings,
              label = rownames(loadings),
              col = "red") +
    geom_segment(data = env_expl,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "blue") +
    geom_text_repel(data = env_expl,
              aes(x = NMDS1, y = NMDS2),
              label = rownames(env_expl),
              col = "blue") +
    geom_point(size = 3, alpha = 0.8) +
    khroma::scale_color_muted() +
    theme_bw() +
    labs(title = "NMDS")

p
}
```

```{r}
plot_nmds2(NERR_Region) + labs(title = "NMDS for all regions", subtitle = "blue = quantitative environmental predictors; \nred = loadings in response matrix")
```

## Bio-env best model graph  


```{r}
env_expl <- env_expl[1:2, ]
```

```{r}
plot_nmds2(NERR_Region) + labs(title = "NMDS for all regions", subtitle = "blue = best quantitative predictor(s) from BIO-ENV; \nred = loadings in response matrix")
```


# Removing Gulf Sites   

```{r}
gulfsites <- which(dat_main$NERR_Region == "Gulf Coast")
responses_scaled <- responses_scaled[-gulfsites, ]
explanatory_scaled <- explanatory_scaled[-gulfsites, ]
explanatory_scaled_main <- explanatory_scaled_main[-gulfsites, ]
```



```{r}
test_bioenv <- bioenv(responses_scaled, explanatory_scaled_main, 
                      method = "spearman",
                      index = "euclidean",
                      metric = "gower")


cat("Explanatory variables used in bioenv: \n", paste("*", names(explanatory_scaled_main)), sep = "\n")

test_bioenv

summary(test_bioenv)
```

## NMDS  

```{r}
veg_mds <- metaMDS(responses_scaled,
                   distance = "euclidean",
                   autotransform = FALSE,
                   trace = FALSE)
cat("2-dimensional stress: ", round(veg_mds$stress, 3))
```

```{r}
data.scores <- as.data.frame(scores(veg_mds))
loadings <- as.data.frame(scores(envfit(veg_mds, responses_scaled), display = "vectors")) *3

# numeric explanatory loadings
expl_numeric <- explanatory_scaled %>% 
    select(SLR_last19yrs,
           condition,
           SET_minus_SLR_19yrs,
           proportion_low)

env_expl <- as.data.frame(scores(envfit(veg_mds, expl_numeric, na.rm = TRUE), display = "vectors")) *5

```



```{r}
# plot all sites
p <- ggplot(data.scores,
            aes(x = NMDS1,
                y = NMDS2)) +
    geom_text(label = rownames(responses_scaled),
              size = rel(3)) +
    geom_segment(data = loadings,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text_repel(data = loadings,
              label = rownames(loadings),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "All sites")
p

# points only
ppts <- ggplot(data.scores,
            aes(x = NMDS1,
                y = NMDS2)) +
    geom_point() +
    geom_segment(data = loadings,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text_repel(data = loadings,
              label = rownames(loadings),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "All sites")
```

## Numeric Loadings overlaid - all  

```{r}
p2 <- ppts +
    geom_segment(data = env_expl,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "blue") +
    geom_text_repel(data = env_expl,
              aes(x = NMDS1, y = NMDS2),
              label = rownames(env_expl),
              col = "blue")

p2
```


## Color by categorical predictors that were important in bioenv  

```{r}
nmds_and_expl <- cbind(data.scores, explanatory_scaled_main)
```

```{r}
plot_nmds2(Salinity_category) + labs(title = "NMDS excluding Gulf Sites", subtitle = "blue = quantitative environmental predictors; \nred = loadings in response matrix")

plot_nmds2(NERR_Region) + labs(title = "NMDS excluding Gulf Sites", subtitle = "blue = quantitative environmental predictors; \nred = loadings in response matrix")
```

## Bio-env best model graph  


```{r}
env_expl <- env_expl[4, ]
```

```{r}
plot_nmds2(Salinity_category) + labs(title = "NMDS excluding Gulf Sites", subtitle = "blue = best quantitative predictor(s) from BIO-ENV; \nred = loadings in response matrix")

plot_nmds2(NERR_Region) + labs(title = "NMDS excluding Gulf Sites", subtitle = "blue = best quantitative predictor(s) from BIO-ENV; \nred = loadings in response matrix")
```


