---
title: "Namaste National Analysis - Ordinations"
output-file: "2024-07-30 NMDS Slopes Veg Groups by Site"
date: today
format: 
    pdf:
        toc: true
    html:
        toc: true
        toc-location: left
echo: false
embed-resources: true
warning: false
message: false
error: true
---

Plot scalings are weird in this file. Living with it for now but can spruce up if we want to include in reports or publications.

```{r}
library(tidyverse)
library(vegan)
library(ggfortify)
library(fastDummies)
```

Using the slopes-by-site data frame.  

```{r}
dat <- read.csv(here::here("data",
                           "compiled",
                           "slopesAndExpl_bySite.csv"))

dat_main <- dat %>% 
    select(Reserve,
           SiteID,
           Total.unvegetated_slope,
           EIR_slope,
           B.Brackish_slope,
           F.Freshwater_slope,
           H.Halophyte_slope,
           U.Upland_slope,
           tide_range_slope:Region,
           condition = NERRs_Landscape_resiliency_condition_sum_quantile,
           SET_change:last_col()
           )

responses <- dat_main %>% 
    select(Total.unvegetated_slope:U.Upland_slope)
row.names(responses) = paste(dat_main$Reserve, dat_main$SiteID, sep = "-")

responses_scaled <- responses %>% 
    mutate(across(everything(), function(x) scale(x)[,1]))

explanatory <- dat_main[which(!(names(dat_main) %in% names(responses)))]
```

is PCA appropriate?  

```{r}
responses_scaled %>%
    mutate(site_num = row_number()) %>% 
    pivot_longer(-site_num,
                 names_to = "response",
                 values_to = "normalized_value") %>% 
    ggplot(aes(x = normalized_value,
               fill = response)) +
    geom_histogram(col = "gray20") +
    facet_wrap(~response) +
    theme(legend.position = "none")
```

Ehhhhh, not great. We also don't have perfect completeness in the dataset, so using NMDS here.



# NMDS  

```{r}
veg_mds <- metaMDS(responses_scaled,
                   distance = "euclidean",
                   autotransform = FALSE,
                   trace = FALSE,
                   na.rm = TRUE,
                   k = 4)
veg_mds$stress
```

Takes 4 axes to get stress below 0.1. With 2 axes, stress is 0.15.

```{r}
data.scores <- as.data.frame(scores(veg_mds)/sqrt(nrow(responses_scaled)))
loadings <- as.data.frame(scores(envfit(veg_mds, responses_scaled, na.rm = TRUE),
                                 "vectors"))
ggplot(data.scores,
       aes(x = NMDS1,
           y = NMDS2)) +
    geom_text(label = rownames(responses_scaled),
              size = rel(3)) +
    geom_segment(data = loadings,
                 aes(x = 0, y = 0,
                  xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text(data = loadings,
              label = rownames(loadings),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "All sites")
```



Again, have Gulf outliers - what happens if we take out the Gulf sites?  

# NMDS v2  


## Non-Gulf only  

```{r}
dat_main_non.gulf <- dat_main %>% 
    filter(!(Reserve %in% c("APA-LSM", "APA-PC", "GND", "MAR")))

responses_non.gulf <- dat_main_non.gulf %>% 
    select(Total.unvegetated_slope:U.Upland_slope)
row.names(responses_non.gulf) = paste(dat_main_non.gulf$Reserve, dat_main_non.gulf$SiteID, sep = "-")

responses_scaled_non.gulf <- responses_non.gulf %>% 
    mutate(across(everything(), function(x) scale(x)[,1]))

explanatory_non.gulf <- dat_main_non.gulf[which(!(names(dat_main_non.gulf) %in% names(responses_non.gulf)))]
```

```{r}
veg_mds_non.gulf <-  metaMDS(responses_scaled_non.gulf,
                   distance = "euclidean",
                   autotransform = FALSE,
                   trace = FALSE,
                   na.rm = TRUE,
                   k = 4)
veg_mds_non.gulf$stress

data.scores_non.gulf <- as.data.frame(scores(veg_mds_non.gulf)/sqrt(nrow(responses_scaled_non.gulf)))

loadings_non.gulf <- as.data.frame(scores(envfit(veg_mds_non.gulf, responses_scaled_non.gulf, na.rm = TRUE), "vectors"))

ggplot(data.scores_non.gulf,
       aes(x = NMDS1,
           y = NMDS2)) +
    geom_text(label = rownames(responses_scaled_non.gulf),
              size = rel(3)) +
    geom_segment(data = loadings_non.gulf,
                 aes(x = 0, y = 0,
                  xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text(data = loadings_non.gulf,
              label = rownames(loadings_non.gulf),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "Non-Gulf sites only")
```


## with add'l loadings (envfit)  

Using main explanatory variables that are going into LME/SEM portion.  

```{r}
main_expl <- c("Geomorphology", "Tidal_Range", "Salinity_category",
               "SLR_last19yrs", "NERR_Region", "condition",
               "SET_minus_SLR_19yrs", "proportion_low")


# select and scale explanatory vars
explanatory_sub <- explanatory %>% 
    select(all_of(main_expl)) %>% 
    mutate(across(c(SLR_last19yrs, condition, SET_minus_SLR_19yrs, proportion_low), scale))

# turn categorical ones into dummy columns
explanatory_sub <- dummy_cols(explanatory_sub, remove_first_dummy = FALSE,
                               remove_selected_columns = TRUE)

# split into numeric and categorical data frames
# for different colored plotting
expl_sub_numeric <- explanatory_sub %>% 
    select(SLR_last19yrs:proportion_low)
expl_sub_cat <- explanatory_sub[ , which(!(names(explanatory_sub) %in% names(expl_sub_numeric)))]
# put it on a similar sd scale as categorical vars by multiplying by 2
expl_sub_cat <- expl_sub_cat*2


# adapt those for non-gulf sites
# select and scale explanatory vars
explanatory_sub <- explanatory_non.gulf %>% 
    select(all_of(main_expl)) %>% 
    mutate(across(c(SLR_last19yrs, condition, SET_minus_SLR_19yrs, proportion_low), scale))

# turn categorical ones into dummy columns
explanatory_sub <- dummy_cols(explanatory_sub, remove_first_dummy = FALSE,
                               remove_selected_columns = TRUE)

# split into numeric and categorical data frames
# for different colored plotting
expl_sub_numeric_non.gulf <- explanatory_sub %>% 
    select(SLR_last19yrs:proportion_low)
expl_sub_cat_non.gulf <- explanatory_sub[ , which(!(names(explanatory_sub) %in% names(expl_sub_numeric_non.gulf)))]
expl_sub_cat_non.gulf <- expl_sub_cat_non.gulf*2


# create separate envfits and data frames of coords for each type
# all sites
# numeric
addl_num <- envfit(veg_mds, expl_sub_numeric, na.rm = TRUE)
vector_num <- as.data.frame(scores(addl_num, "vectors"))
mult_num <- ordiArrowMul(addl_num)  # multiplier for scaling plots
# categorical
addl_cat <- envfit(veg_mds, expl_sub_cat, na.rm = TRUE)
vector_cat <- as.data.frame(scores(addl_cat, "vectors"))

# non-gulf
# numeric
addl_num_non.gulf <- envfit(veg_mds_non.gulf, expl_sub_numeric_non.gulf, na.rm = TRUE)
vector_num_non.gulf <- as.data.frame(scores(addl_num_non.gulf, "vectors"))
mult_num_non.gulf <- ordiArrowMul(addl_num_non.gulf)  # multiplier for scaling plots
# categorical
addl_cat_non.gulf <- envfit(veg_mds_non.gulf, expl_sub_cat_non.gulf, na.rm = TRUE)
vector_cat_non.gulf <- as.data.frame(scores(addl_cat_non.gulf, "vectors"))
```


# Plots - All Sites  

## Only Responses overlaid  

```{r}
# plot all sites
p <- ggplot(data.scores,
            aes(x = NMDS1,
                y = NMDS2)) +
    geom_text(label = rownames(responses_scaled),
              size = rel(3)) +
    geom_segment(data = loadings,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text(data = loadings,
              label = rownames(loadings),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "All sites")
p

# points only
ppts <- ggplot(data.scores,
            aes(x = NMDS1,
                y = NMDS2)) +
    geom_point() +
    geom_segment(data = loadings,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text(data = loadings,
              label = rownames(loadings),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "All sites")
```


## Numeric Loadings overlaid  

```{r}
p2 <- ppts +
    geom_segment(data = vector_num,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "blue") +
    geom_text(data = vector_num,
              aes(x = NMDS1, y = NMDS2),
              label = rownames(vector_num),
              col = "blue")

p2
```



## Categorical Loadings overlaid  

```{r}
p3 <- ppts +
    geom_segment(data = vector_cat,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "purple") +
    geom_text(data = vector_cat,
              aes(x = NMDS1, y = NMDS2),
              label = rownames(vector_cat),
              col = "purple",
              size = rel(3))
p3
```




## Color by categories instead  

```{r}
# divide by the square root of number of rows
# to get graph scaling factor
nmds_and_expl <- cbind(scores(veg_mds)/sqrt(nrow(explanatory)), explanatory)
nmds_and_expl_non.gulf <- cbind(scores(veg_mds_non.gulf)/sqrt(nrow(explanatory_non.gulf)), explanatory_non.gulf)
```


```{r}
p <- ggplot(data.scores,
            aes(x = NMDS1,
                y = NMDS2)) +
    geom_segment(data = loadings,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text(data = loadings,
              label = rownames(loadings),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "All sites") 

num_loadings <- geom_segment(data = vector_num,
                             aes(x = 0, y = 0,
                                 xend = NMDS1, yend = NMDS2),
                             col = "blue") 
num_labels <- geom_text(data = vector_num,
                        aes(x = NMDS1, y = NMDS2),
                        label = rownames(vector_num),
                        col = "blue")

p +
    geom_point(data = nmds_and_expl,
               aes(x = NMDS1,
                   y = NMDS2,
                   col = NERR_Region),
               size = 3, alpha = 0.8) +
    khroma::scale_color_muted() +
    labs(subtitle = "Colored by Region") +
    num_loadings +
    num_labels
    


p +
    geom_point(data = nmds_and_expl,
               aes(x = NMDS1,
                   y = NMDS2,
                   col = Geomorphology),
               size = 3, alpha = 0.8) +
    khroma::scale_color_muted() +
    labs(subtitle = "Colored by Geomorphology") +
    num_loadings +
    num_labels

p +
    geom_point(data = nmds_and_expl,
               aes(x = NMDS1,
                   y = NMDS2,
                   col = Salinity_category),
               size = 3, alpha = 0.8) +
    khroma::scale_color_muted() +
    labs(subtitle = "Colored by Salinity Category") +
    num_loadings +
    num_labels


p +
    geom_point(data = nmds_and_expl,
               aes(x = NMDS1,
                   y = NMDS2,
                   col = Tidal_Range),
               size = 3, alpha = 0.8) +
    khroma::scale_color_muted() +
    labs(subtitle = "Colored by Tidal range") +
    num_loadings +
    num_labels
```




# Plots - Non-Gulf Sites  

## Only Responses overlaid  

```{r}
# plot all sites
p <- ggplot(data.scores_non.gulf,
            aes(x = NMDS1,
                y = NMDS2)) +
    geom_text(label = rownames(responses_scaled_non.gulf),
              size = rel(3)) +
    geom_segment(data = loadings_non.gulf,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text(data = loadings_non.gulf,
              label = rownames(loadings_non.gulf),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "All sites")
p

# points only
ppts <- ggplot(data.scores_non.gulf,
            aes(x = NMDS1,
                y = NMDS2)) +
    geom_point() +
    geom_segment(data = loadings_non.gulf,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text(data = loadings_non.gulf,
              label = rownames(loadings_non.gulf),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "All sites")
```


## Numeric Loadings overlaid  

```{r}
p2 <- ppts +
    geom_segment(data = vector_num_non.gulf,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "blue") +
    geom_text(data = vector_num_non.gulf,
              aes(x = NMDS1, y = NMDS2),
              label = rownames(vector_num_non.gulf),
              col = "blue")

p2
```



## Categorical Loadings overlaid  

```{r}
p3 <- ppts +
    geom_segment(data = vector_cat_non.gulf,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "purple") +
    geom_text(data = vector_cat_non.gulf,
              aes(x = NMDS1, y = NMDS2),
              label = rownames(vector_cat_non.gulf),
              col = "purple",
              size = rel(3))
p3
```




## Color by categories instead  


```{r}
p <- ggplot(data.scores_non.gulf,
            aes(x = NMDS1,
                y = NMDS2)) +
    geom_segment(data = loadings_non.gulf,
                 aes(x = 0, y = 0,
                     xend = NMDS1, yend = NMDS2),
                 col = "red") +
    geom_text(data = loadings_non.gulf,
              label = rownames(loadings_non.gulf),
              col = "red") +
    labs(title = "NMDS",
         subtitle = "Non-Gulf only sites") 

num_loadings <- geom_segment(data = vector_num_non.gulf,
                             aes(x = 0, y = 0,
                                 xend = NMDS1, yend = NMDS2),
                             col = "blue") 
num_labels <- geom_text(data = vector_num_non.gulf,
                        aes(x = NMDS1, y = NMDS2),
                        label = rownames(vector_num_non.gulf),
                        col = "blue")

p +
    geom_point(data = nmds_and_expl_non.gulf,
               aes(x = NMDS1,
                   y = NMDS2,
                   col = NERR_Region),
               size = 3, alpha = 0.8) +
    khroma::scale_color_muted() +
    labs(subtitle = "Colored by Region") +
    num_loadings +
    num_labels
    


p +
    geom_point(data = nmds_and_expl_non.gulf,
               aes(x = NMDS1,
                   y = NMDS2,
                   col = Geomorphology),
               size = 3, alpha = 0.8) +
    khroma::scale_color_muted() +
    labs(subtitle = "Colored by Geomorphology") +
    num_loadings +
    num_labels

p +
    geom_point(data = nmds_and_expl_non.gulf,
               aes(x = NMDS1,
                   y = NMDS2,
                   col = Salinity_category),
               size = 3, alpha = 0.8) +
    khroma::scale_color_muted() +
    labs(subtitle = "Colored by Salinity Category") +
    num_loadings +
    num_labels


p +
    geom_point(data = nmds_and_expl_non.gulf,
               aes(x = NMDS1,
                   y = NMDS2,
                   col = Tidal_Range),
               size = 3, alpha = 0.8) +
    khroma::scale_color_muted() +
    labs(subtitle = "Colored by Tidal range") +
    num_loadings +
    num_labels
```




# Other  

## Condition?  

```{r}
ggplot(dat_main, aes(x = Reserve, y = condition)) + geom_jitter() + theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

