---
title: "Untitled"
format: 
    html:
        toc: true
        toc-location: left
        code-fold: true
        embed-resources: true
warning: false
message: false
error: true
---

```{r}
library(tidyverse)
```

```{r}
slopes <- read.csv(here::here("data",
                              "compiled",
                              "slopesAndExpl_byPlot.csv"))
slopes2 <- read.csv(here::here("data",
                              "compiled",
                              "slopesAndExpl_bySite.csv"))
```

Potential responses - change in:  

-  Total unvegetated OR total live vegetated
-  EIR  
-  Richness  
-  Diversity  
-  Salt-to-Total - not a lot of variation in this though    
-  Halophyte cover  
-  Brackish cover  

Potential predictors that we have for all plots:  

-  tide range slope  
-  temperature slope  
-  precip slope  
-  geomorphology (categorical)    
-  tidal range (categorical)  
-  salinity category (categorical)  
-  SLR since 1970  
-  SLR last 19 years  
-  Latitude  
-  Region  

Potential predictors that we don't have for all plots (proportion we have is in parentheses; sorted from highest to lowest):  

-  Landscape resiliency (0.85) .... one or more of the metrics? condition? vulnerability? Pct MUC below MHHW?  
-  distance to water (0.75 at plot level)  
-  SET change (0.69)  
-  SET deficit (0.69; careful of collinearity with SLR metrics)  
-  orthometric height (0.67 at plot level)  
-  MARS score (0.58)  
 

# Exploring possible responses  

Need a good spread of values in the response.  

```{r}
responses_long <- slopes %>% 
    mutate(plot_full = paste(Reserve, SiteID, TransectID, PlotID, sep = "-")) %>% 
    select(plot_full,
           Total.unvegetated_slope,
           Total.live.veg_slope,
           H.Halophyte_slope,
           B.Brackish_slope,
           EIR_slope,
           Richness_slope,
           SWdiv_slope,
           Salt_to_Total_slope) %>% 
    pivot_longer(-plot_full,
                 names_to = "response",
                 values_to = "slope")
```

```{r}
ggplot(responses_long,
       aes(x = slope,
           fill = response)) +
    geom_histogram(col = "gray40") + 
    geom_vline(xintercept = 0,
               col = "black") +
    facet_wrap(~response, scales = "free") +
    khroma::scale_fill_muted() +
    theme_bw() +
    theme(legend.position = "none")
```

Looks like total live/unvegetated are both fine, as is EIR, Halophytes, richness, and diversity. Noting though that increasing richness isn't necessarily always a good thing (the species coming in can be invasive). So we might want to focus on live/unvegetated, EIR, and Halophytes.  

Also, just summarizing these slopes is probably useful - median change around 0 or not? Where are the quantiles? Etc. (after figuring out what the independent units are, so we don't over-represent reserves or regions that have more sampling plots than others)  