---
title: "Ordinations"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      error = TRUE)
```

```{r}
library(tidyverse)
library(vegan)
library(lme4)
library(emmeans)
```


Exploring Ordination possibilities to see whether the analysis plan will work.  


There are definitely some outlier values in this dataset. Those may go away when we reduce according to the flow chart in the analysis plan - should do that and then check again.  

```{r}
load(here::here("data", "compiled", "slopesAndExpl_dfs.RData"))
```

# By Plot  

```{r}
plot_forOrd <- slopesAndExpl_byPlot %>% 
    mutate(RsStTrPl = paste(Reserve, SiteID, TransectID, PlotID,
                            sep = "_")) %>% 
    column_to_rownames("RsStTrPl")

plot_resp <- plot_forOrd %>% 
    select(Total.unvegetated_slope:Wood_slope)

plot_expl <- plot_forOrd %>% 
    select(-any_of(names(plot_resp)))
```

Do we have generally normal slopes, which would be great for PCA? What do our response slopes look like?  

```{r}
skimr::skim(plot_resp)
```


```{r}
plot_long <- plot_resp %>% 
    rownames_to_column("RsStTrPl") %>% 
    pivot_longer(-RsStTrPl,
                 names_to = "response",
                 values_to = "value")

ggplot(plot_long, aes(x = value,
                      fill = response)) +
    geom_histogram(bins = 30, col = "gray40") +
    scale_x_log10() +
    geom_vline(xintercept = 1e-13) +
    facet_wrap(~response, nrow = 5) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(subtitle = "NOTE log-10 x-axis")
```

With a log transformation it wouldn't look too bad, but there are some pretty giant outliers.  


# By Zone (within site)  



# By Site  

```{r}
site_forOrd <- slopesAndExpl_bySite %>% 
    mutate(RsSt = paste(Reserve, SiteID,
                            sep = "_")) %>% 
    column_to_rownames("RsSt")

site_resp <- site_forOrd %>% 
    select(Total.unvegetated_slope:Wood_slope)

site_expl <- site_forOrd %>% 
    select(-any_of(names(site_resp)))

skimr::skim(site_resp)

site_long <- site_resp %>% 
    rownames_to_column("RsSt") %>% 
    pivot_longer(-RsSt,
                 names_to = "response",
                 values_to = "value")

ggplot(site_long, aes(x = value,
                      fill = response)) +
    geom_histogram(bins = 30, col = "gray40") +
    scale_x_log10() +
    geom_vline(xintercept = 1e-13) +
    facet_wrap(~response, nrow = 5) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(subtitle = "NOTE log-10 x-axis")
```

# Slopes ~ explanatory vars  

## plot-level  

```{r}
# probably need to do this for each vegetation zone
# which means I need zone in the data frame
plot_forLMM <- plot_forOrd %>% 
    select(Reserve:PlotID,
           Total.live.veg_slope,
           Total.unvegetated_slope,
           H.Halophyte_slope,
           EIR_slope,
           Richness_slope,
           SWdiv_slope,
           Unveg_to_veg_slope,
           Geomorphology,
           Tidal_Range,
           Salinity_category,
           NERR_Region,
           Region) %>% 
    mutate(Plot_full = paste(TransectID, PlotID, sep = "_"))

preds <- paste(names(plot_forLMM)[12:15], collapse = " + ")
form_rs <- paste(preds, " + (1|Reserve)")

fo <- paste0("Total.live.veg_slope ~ ", form_rs)
mod <- lmer(as.formula(fo),
            data = plot_forLMM)

emmeans(mod, ~NERR_Region)
emmeans(mod, ~Geomorphology)
emmeans(mod, ~Tidal_Range)
emmeans(mod, ~Salinity_category, by = "NERR_Region") # would really want this to be by = "Vegetation_Zone" - it splits the tables up nicely
```

## zone (within site) level  

```{r}
# probably need to do this for each vegetation zone
# which means I need zone in the data frame
zone_forLMM <- slopesAndExpl_byZone %>% 
    select(Reserve:Vegetation_Zone,
           Total.live.veg_slope,
           Total.unvegetated_slope,
           H.Halophyte_slope,
           EIR_slope,
           Richness_slope,
           SWdiv_slope,
           Unveg_to_veg_slope,
           Geomorphology,
           Tidal_Range,
           Salinity_category,
           NERR_Region,
           Region) 

preds <- paste(names(zone_forLMM)[11:14], collapse = " + ")
form_rs <- paste(preds, " + Vegetation_Zone + (1|Reserve)")

fo <- paste0("EIR_slope ~ ", form_rs)
mod <- lmer(as.formula(fo),
            data = zone_forLMM)

emmeans(mod, ~NERR_Region, by = "Vegetation_Zone")
emmeans(mod, ~Geomorphology, by = "Vegetation_Zone")
emmeans(mod, ~Tidal_Range, by = "Vegetation_Zone")
emmeans(mod, ~Salinity_category, by = "Vegetation_Zone") 
emmeans(mod, ~NERR_Region*Salinity_category, by = "Vegetation_Zone") 

emmip(mod, ~NERR_Region*Vegetation_Zone,
      CIs = TRUE)
emdf1 <- emmip(mod, ~NERR_Region*Vegetation_Zone,
      CIs = TRUE, plotit = FALSE)
ggplot(emdf1, aes(x = NERR_Region,
                  y = yvar,
                  col = NERR_Region)) +
    geom_hline(yintercept = 0,
               col = "gray40") +
    geom_pointrange(aes(ymin = LCL,
                        ymax = UCL),
                    linewidth = 0.8) +
    facet_wrap(~Vegetation_Zone, ncol = 1) +
    theme_bw() +
    scale_color_brewer(palette = "Set1") +
    theme(legend.position = "none") +
    labs(title = "EIR slopes by vegetation zone",
         y = "change in EIR/year",
         x = "Region")

# what about throwing that on top of the raw data????
ggplot() +
    geom_jitter(data = zone_forLMM,
                aes(x = NERR_Region,
                    y = EIR_slope),
                color = "gray") +
    geom_hline(yintercept = 0) +
    geom_pointrange(data = emdf1,
                    aes(x = NERR_Region,
                        y = yvar,
                        ymin = LCL,
                        ymax = UCL,
                        col = NERR_Region),
                    linewidth = 0.8) +
    facet_wrap(~Vegetation_Zone, ncol = 1) +
    theme_bw() +
    scale_color_brewer(palette = "Set1") +
    theme(legend.position = "none") +
    labs(title = "EIR slopes by vegetation zone",
         y = "change in EIR/year",
         x = "Region")

# should we just do a loop and get average and SE/CIs of each slope by zone*whatever other explanatory var???
```